Nonplanarity and Street Network Representation in Urban Form Studies

### 2018-01-24 manuscript.tex
edits

Mathematical models of street networks, called graphs, are widely used throughout the urban studies literature to examine travel behavior, accessibility, urban design patterns, and morphology. These networks are commonly called planar graphs, meaning that they can be represented in two dimensions without any underpasses or overpasses. However, real urban street networks exist in three-dimensional space and frequently include grade separation such as bridges and tunnels. How well do planar graphs model real-world street networks? What does the extent of nonplanarity tell us about a place's infrastructure and urban development? We measure the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examine the variation of planarity across a single city. We find that while some street networks are approximately planar, many are poorly-modeled by planar graphs. In particular, planarity can drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths.

@@

In urban planning and design research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity. Mathematical models of street networks, called graphs, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns and spatial morphology, and location centrality and polycentricity \citep{southworth_street_1995,marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.

@@

Consider the examples in Figure \ref{fig:planar_vs_not}. The left network comprises a set of paths and streets without any bridges or tunnels. In two-dimensions, its edges intersect each other only at nodes so it is by definition planar. The right network comprises surface streets and a grade-separated freeway interchange with many overpasses and underpasses. In two-dimensions, its edges sometimes cross each other at non-nodes (i.e., overpasses), so it is spatially nonplanar. Furthermore it is impossible to redraw the graph by moving its nodes and edges around in space such that its edges only intersect at nodes. Therefore it is also formally nonplanar. However, at a city- or region-wide spatial scale, these nonplanar edge crossings may be relatively uncommon; we might call such a network \emph{approximately} planar. Approximate planarity constrains nonplanar spatial networks such that they do not exhibit certain characteristics found among nonplanar aspatial graphs, such as small-world effects or power-law distributed node degrees \citep{fischer_spatial_2014}.

@@

OpenStreetMap's raw data contain many interstitial nodes in the middle of street segments (forming an expansion graph) to allow streets to curve through space via a series of straight-line approximations. OSMnx automatically simplifies each graph's topology to retain nodes only at intersections and dead-ends, while faithfully retaining each edge's true spatial geometry. This provides us with an accurate count of intersections and an accurate measure of edge lengths for comparison between the planar and nonplanar representations of our networks.

@@

This indicates the extent to which planarity overstates intersections and connectivity in each street network. In other words, $\phi$ tells us what proportion of the two-dimensional edge crossings in the planar graph are true intersections in the nonplanar graph. A truly planar network with no bridges or tunnels will thus have an $\phi$ score of 1.0, while lower values indicate the percentage extent to which the network is planar. From the $\phi$, we can further calculate by what percentage the planar graph overstates intersection counts as:

@@

Finally, we explore how these three indicators vary across a single city. To do so we analyze the drivable street network of Oakland, California, a reasonably representative midsized American city with a variety of urban form types from gridded street patterns in its flatlands, to winding culs-de-sac in its hills, to freeways and dense blocks around its downtown. First we analyze the entire city of Oakland. Then we recreate the aforementioned methodology by randomly sampling 100 points within the city limits and analyzing the square-mile street networks centered on each. The resulting statistical dispersion of planarity demonstrates the extent to which analyzing an entire city's neighborhoods as a single graph may obscure neighborhood-scale planarity characteristics.

@@

Table \ref{tab:world_cities} shows $\phi$, $\lambda$, and results of the formal planarity test for the walkable and the drivable networks in these 50 city centers. Among the drivable street networks, only 20\% are formally planar. On average, they are 88.5\% planar by the $\phi$ measure and 90\% planar by the $\lambda$ measure. The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 57\% in Moscow. The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 63.5\% in Los Angeles. On average across these networks, planar representations overcount intersections by 16\% and underestimate street segment lengths by 10\%.

@@

Not all formally planar street networks are spatially planar. For example, Toronto's drivable network is formally planar but only 93\% and 96\% spatially planar according to its $\phi$ and $\lambda$ measures. In total, three drivable networks (Toronto, Jakarta, and Copenhagen) and three walkable networks (Dallas, Delhi, and Bologna) are formally planar but spatially nonplanar to various extents. About a third of the cities studied demonstrate spatial planarity of 97\% or higher according to the $\phi$, suggesting that they are \enquote{approximately} planar. However, another third of the cities studied are less than 87\% planar. Dallas, Los Angeles, and Moscow have $\phi$ values below 60\%, suggesting planar graphs poorly model these city centers. Moreover, planar graphs overstate the intersection counts in these three networks by 67\%, 72\%, and 74\% respectively (Equation \ref{eq:spr_overstates}).

Mogadishu is the only city studied that demonstrates perfect planarity across all three indicators for both network types. All three Italian cities demonstrate perfect planarity in their centers' drivable networks, but not in their walkable networks. The extent of planarity is not consistent across network types: Dallas's walkable $\phi$ is 61\% greater than its drivable $\phi$, while Geneva's drivable $\phi$ is 19\% greater than its walkable $\phi$. Figure \ref{fig:world_map_bw} shows the distribution of $\phi$ values around the world. While nearly every European city is in the highest tercile, indicating their networks are more planar, most American cities are in the lowest tercile, indicating their networks are more nonplanar.

Figure \ref{fig:regression_split} shows the relationship between the $\phi$ and $\lambda$ indicators across all 50 cities for both network types, as a log-log plot. The indicators' log-log relationship is linear, positive, and very strong (drivable $r^2=0.98$ and walkable $r^2=0.99$). This strong correlation tells us that these two indicators unsurprisingly provide redundant statistical information about the extent to which a network is planar. However, each assesses different implications of this bias for measuring the urban form.

@@

Finally, we examine how these measures behave across an entire city. We find that Oakland's city-wide street network is formally nonplanar. The city has a $\phi$ score of 91.8\% and an $\lambda$ score of 93.6\%. This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections --- and thus, the network's connectivity --- by 8.9\% city-wide (Equation \ref{eq:spr_overstates}) and understates the average edge length by 6.4\% city-wide (Equation \ref{eq:elr_understates}).

However, these indicators' values vary across the city. To explore this statistical variation, Table \ref{tab:samples_city} presents summary statistics of these planarity indicators across 100 square-mile samples of Oakland's drivable street network. Our samples' mean $\phi$ and $\lambda$ scores are reasonably close to the city-wide values. However, the samples range from spatial planarity lows of 56.9\% ($\phi$) and 63.7\% ($\lambda$) up to highs of 100\%. 67\% of the samples pass the formal planarity test. However, 63\% of the samples are at least somewhat spatially nonplanar (i.e., with $\phi$ less than 1).

@@

Our findings suggest that the street networks at the centers of most major cities are formally nonplanar. However, this depends on the scale of measurement: across an entire city there is likely to be at least one overpass or underpass somewhere, while individual neighborhoods or small towns might be formally planar in their entirety. The type and era of urbanization represent another factor. Medieval European towns or informal settlements in the global south may contain fewer grade-separated roads --- and thus are more planar --- than 20th-century American or 21st-century Chinese metropolises. This is a result of the prevailing transportation technologies when the urban form was developed, as well as the local terrain, wealth, culture, and politics.

@@

Contrary to some of the statements in the urban studies literature, our results suggest that it cannot be universally claimed that urban street networks are planar graphs. But perhaps some of it comes down to vocabulary. If road is not synonymous with street, then a road network and a street network may not be synonymous. A road network, including freeways and boulevards, may frequently be nonplanar, but a street network, focusing on municipal streets lined by land parcels, may be at-grade and planar. But this ignores the fact that even residential streets sometimes include bridges and tunnels in hilly neighborhoods, and the fact that our analysis earlier showed that walkable circulation networks in city centers often include pedestrian tunnels and footbridges. A graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes. Thus, street networks are not universally planar graphs.

However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is planar.

@@

As George Box once said, \enquote{All models are wrong but some are useful.} Even if they are not formally planar, can street networks be simplified to planar graphs and still be usefully well-modeled? Our results suggest that the answer depends on the study site and on the type of analysis. In limited circumstances, where the circulation network and built form exhibit few (or ideally zero) underpasses, overpasses, or grade-separation, then perhaps yes.

@@

But planar graphs are often used in the literature to characterize urban form and morphology. So, aside from routing, do planar graphs offer \emph{useful} models for this type of research? Again, only in limited circumstances, such as the drivable networks in the three Italian cities we analyzed. Elsewhere, the results in Table \ref{tab:world_cities} showed how common urban form measures such as intersection counts are overstated by planar models, while average street segment lengths are consequently understated. Moreover, this misrepresentation behaves inconsistently from place to place: Figure \ref{fig:world_map_bw} demonstrated how the magnitude of bias varies across cities and modes of urbanization. Although planar graphs offer computationally tractable models and allow for the polygonal spatial analysis of urban blocks, they often modeled these street networks poorly compared to nonplanar graphs.

@@

We might repurpose this discrepancy between planar and nonplanar representations of a street network to reveal something useful about urban form. The $\phi$ and $\lambda$ scores of the major US city centers indicate the greater three-dimensionality of their transportation infrastructure compared to that of the European city centers. These US city centers feature more automobile-oriented roadways and are often bounded by grade-separated freeways. By contrast, these European city centers feature less three-dimensionality as their streets and paths tend to be at-grade. Examining how $\phi$ and $\lambda$ change over time in different cities could tell us something about the type and pace of urban development.

It is also interesting that most street networks are formally nonplanar while demonstrating $\phi$ and $\lambda$ scores that are only slightly nonplanar. Approximate planarity is common across cities because of costs, technology, and politics. Why? It is expensive to engineer a three-dimensional network with z-axis extents (nearly) as extensive as the x- and y-axes. Furthermore, a city fully consumed by overpasses and tunnels would present degraded livability and walkability, making it politically infeasible.

@@

Urban researchers often use planar graphs to model street networks. This study demonstrated that, although in limited circumstances these models may be accurate, they behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths. In most circumstances, nonplanar graphs provide better models. It also demonstrated how these indicators can be used to characterize urbanization in different cities, in particular through the transportation infrastructure's three-dimensionality.

>>>>

Mathematical models of street networks, called graphs, are widely used throughout the urban studies literature to examine travel behavior, accessibility, urban design patterns, and morphology. These graphs are commonly described as \enquote{planar,} meaning that they can be represented in two dimensions without any underpasses or overpasses. However, real urban street networks exist in three-dimensional space and frequently include grade separation such as bridges and tunnels. How well do planar graphs model real-world street networks? What does the extent of nonplanarity tell us about a city's infrastructure and urban development? We measure the nonplanarity of drivable and walkable street networks in the centers of 50 cities worldwide, then examine the variation of planarity across a single city. We find that while some street networks are approximately planar, many are poorly-modeled by planar graphs. In particular, planarity can inconsistently but drastically misrepresent routing, connectivity, intersection counts/densities, and block lengths. We develop two measures of a network's nonplanarity and argue it can be a useful indicator of transportation infrastructure and development.

@@

In urban planning and design research, street networks are routinely used to calculate accessibility between origins and destinations or to compute indicators of the urban form, such as block sizes or intersection density and connectivity. Mathematical models of street networks, called graphs, have grown ubiquitous in the urban studies literature in recent years as they have been used to model household travel patterns, access and equity, pedestrian volume, urban design patterns and spatial morphology, and location centrality and polycentricity \citep{marshall_street_2010,porta_alterations_2014,marshall_community_2014,hajrasouliha_impact_2015,parthasarathi_street_2015,knight_metrics_2015,gil_street_2016,zhong_revealing_2017}.

@@

Consider the examples in Figure \ref{fig:planar_vs_not}. The left network comprises a set of paths and streets without any bridges or tunnels. In two-dimensions, its edges intersect each other only at nodes so it is by definition planar. The right network comprises surface streets and a grade-separated freeway interchange with many overpasses and underpasses. In two-dimensions, its edges sometimes cross each other at non-nodes (i.e., overpasses), so it is spatially nonplanar. Furthermore it is impossible to redraw the graph by moving its nodes and edges around in space such that its edges only intersect at nodes. Therefore it is also formally nonplanar. However, at a city- or region-wide spatial scale, these nonplanar edge crossings may be relatively uncommon; we might call such a network \emph{approximately} planar. Approximate planarity constrains nonplanar spatial networks such that they do not exhibit certain characteristics found among nonplanar aspatial graphs, such as small-world effects or power-law distributed node degrees \citep{crucitti_centrality_2006,fischer_spatial_2014}.

@@

OpenStreetMap's raw data contain many interstitial nodes in the middle of street segments (forming an expansion graph) to allow streets to curve through space via a series of straight-line approximations. OSMnx automatically simplifies each graph's topology to retain nodes only at intersections and dead-ends, while faithfully retaining each edge's true spatial geometry. This provides an accurate count of intersections and an accurate measure of edge lengths for comparison between the planar and nonplanar representations of our networks.

@@

Thus, $\phi$ represents what proportion of the two-dimensional edge crossings in the planar graph are true intersections in the nonplanar graph. This indicates the extent to which planarity overstates intersections and connectivity in a street network. A truly planar network with no bridges or tunnels will have an $\phi$ score of 1.0, while lower values indicate the percentage extent to which the network is planar. From this indicator, we can further calculate by what percentage the planar graph overstates intersection counts as:

@@

Finally, we explore how these three indicators vary across a single city. To do so we analyze the drivable street network of Oakland, California as a case study. Oakland is a reasonably representative midsized American city with a variety of urban form types from gridded street patterns in its flatlands, to winding culs-de-sac in its hills, to freeways and dense blocks around its downtown. First we analyze the entire city of Oakland. Then we recreate the aforementioned methodology by randomly sampling 100 points within the city limits and analyzing the square-mile street networks centered on each. The resulting statistical dispersion of planarity demonstrates the extent to which analyzing an entire city's neighborhoods as a single graph may obscure neighborhood-scale infrastructure characteristics.

@@

Table \ref{tab:world_cities} lists the $\phi$, $\lambda$, and formal planarity of the walkable and drivable networks in these 50 city centers. Among the drivable street networks, only 20\% are formally planar. On average, they are 88.5\% planar by the $\phi$ measure and 90\% planar by the $\lambda$ measure. The individual $\phi$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 57\% in Moscow. The $\lambda$ values indicate that spatial planarity ranges from a high of 100\% in seven of these cities to a low of 63.5\% in Los Angeles. On average across these networks, planar representations overcount intersections by 16\% and underestimate street segment lengths by 10\% (Equations \ref{eq:spr_overstates} and \ref{eq:elr_understates}).

@@

Not all formally planar street networks are spatially planar. For example, Toronto's drivable network is formally planar but only 93\% ($\phi$) and 96\% ($\lambda$) spatially planar. In total, three drivable networks (Toronto, Jakarta, and Copenhagen) and three walkable networks (Dallas, Delhi, and Bologna) are formally planar but spatially nonplanar to various extents. About a third of the cities studied demonstrate $\phi$ spatial planarity of 97\% or higher, suggesting that they are \enquote{approximately} planar. However, another third of the cities studied are less than 87\% planar. Dallas, Los Angeles, and Moscow have $\phi$ values below 60\%, suggesting planar graphs poorly model these city centers. Moreover, planar graphs overstate the intersection counts in these three networks by 67\%, 72\%, and 74\% respectively.

Mogadishu is the only city studied that demonstrates perfect planarity across all three indicators for both network types. All three Italian cities demonstrate perfect planarity in their centers' drivable networks, but not in their walkable networks. The extent of planarity is not consistent across network types: Dallas's walkable $\phi$ is 61\% greater than its drivable $\phi$, while Geneva's drivable $\phi$ is 19\% greater than its walkable $\phi$. Figure \ref{fig:world_map_bw} maps the distribution of $\phi$ values around the world. While nearly every European city is in the highest tercile, indicating their networks are more planar, most American cities are in the lowest tercile, indicating their networks are more nonplanar.

Figure \ref{fig:regression_split} depicts the relationship between the $\phi$ and $\lambda$ indicators across all 50 cities for both network types, as a log-log plot. The indicators' log-log relationship is linear, positive, and very strong (drivable $r^2=0.98$ and walkable $r^2=0.99$). The coefficients of determination tell us that these two indicators unsurprisingly provide redundant statistical information about the extent to which a network is planar. However, each assesses different implications of this bias for measuring the urban form.

@@

Finally, we examine how these measures behave across an entire city. We find that Oakland's city-wide street network is formally nonplanar. The city has a $\phi$ score of 91.8\% and an $\lambda$ score of 93.6\%. This suggests that the planar representation of Oakland's drivable street network overstates the number of intersections --- and thus, the network's connectivity --- by 8.9\% city-wide and understates the average edge length by 6.4\% city-wide.

However, these indicators' values vary across the city. To explore this statistical variation, Table \ref{tab:samples_city} presents summary statistics of these planarity indicators across 100 square-mile samples of Oakland's drivable street network. Our samples' mean $\phi$ and $\lambda$ scores are reasonably close to the city-wide values. However, the samples range from spatial planarity lows of 56.9\% ($\phi$) and 63.7\% ($\lambda$) up to highs of 100\%. 67\% of the samples pass the formal planarity test; however, 63\% of the samples are at least somewhat spatially nonplanar (i.e., with $\phi < 1$).

@@

Our findings suggest that the street networks at the centers of most major cities are formally nonplanar. However, this depends on the scale of measurement: across an entire city there is likely to be at least one overpass or underpass somewhere, while individual neighborhoods or small towns might be formally planar in their entirety. The type and era of urbanization represent another factor. Medieval European towns or informal settlements in the global south may contain fewer grade-separated roads --- and thus are more planar --- than 20th-century American or 21st-century Chinese metropolises. This is a result of the prevailing transportation technologies when the urban form was developed, as well as local terrain, wealth, culture, and politics \citep{southworth_street_1995}.

@@

Contrary to some of the statements in the urban studies literature, our results suggest that it cannot be universally claimed that urban street networks are planar graphs. A graph is not planar because its edges \emph{usually} intersect only at nodes: by definition it is planar because its edges \emph{exclusively} intersect at nodes. However, as \citet{newman_networks:_2010} points out, debating the semantics of formal planarity may be missing the point -- more interesting is the \emph{extent} to which a network is planar.

@@

As George Box famously said, \enquote{All models are wrong but some are useful.} Even if they are not formally planar, can street networks be simplified to planar graphs and still be usefully well-modeled? Our results suggest that the answer depends on the study site and on the type of analysis. In limited circumstances --- where the circulation network exhibits few (or ideally zero) underpasses, overpasses, or grade-separation --- perhaps yes.

@@

But planar graphs are often used in the literature to characterize urban form and morphology. So, aside from routing, do planar graphs offer \emph{useful} models for this type of research? Again, only in limited circumstances, such as the drivable networks in the three Italian cities we analyzed. Elsewhere, the results in Table \ref{tab:world_cities} showed how common urban form measures such as intersection counts are overstated by planar models (16\% on average in the drivable networks), while average street segment lengths are consequently understated (10\% on average in the drivable networks). Moreover, this misrepresentation behaves inconsistently from place to place: Figure \ref{fig:world_map_bw} and Table \ref{tab:samples_city} demonstrated how the magnitude of bias varies across cities and modes of urbanization. Although planar graphs offer computationally tractable models and allow for the polygonal spatial analysis of urban blocks, they often model these street networks poorly when compared to nonplanar graphs.

@@

We might repurpose this discrepancy between planar and nonplanar representations of a street network to reveal something useful about urban form. The $\phi$ and $\lambda$ scores of the major US city centers indicate the greater three-dimensionality of their transportation infrastructure compared to that of the European city centers. These US city centers feature more automobile-oriented roadways and are often bounded by grade-separated freeways. By contrast, these European city centers feature less three-dimensionality as their streets and paths tend to be at-grade. Examining how $\phi$ and $\lambda$ change over time in different cities could reveal the type and pace of urban development.

Most street networks are formally nonplanar while demonstrating $\phi$ and $\lambda$ values that are only slightly nonplanar. Approximate planarity is common across cities because of costs, technology, and politics: it is expensive to engineer a three-dimensional network with z-axis extents (nearly) as extensive as the x- and y-axes. Furthermore, a city fully consumed by overpasses and tunnels would degrade livability and walkability, making it politically infeasible.

@@

Urban researchers often use planar graphs to model street networks. This study demonstrated that, although in limited circumstances these models may be accurate, they behave inconsistently across different kinds of cities by misrepresenting connectivity, accessibility, routing, intersection counts and densities, and street segment lengths. In most circumstances, nonplanar graphs provide more faithful models. It also demonstrated how these indicators can characterize urbanization in different cities, particularly through transportation infrastructure's three-dimensionality.
